---
title: "DATA VISUALIZATION"
output:
  html_document:
    toc: true
    toc_depth: 2

bibliography: bib/thesis.bib
biblio-style: apalike
link-citations: yes
---

```{r setup, include=FALSE}
# load libraries
library(knitr)
library(data.table)
library(tidyverse)
library(kableExtra)
library(data.table)
library(tidyverse)
library(plotly)
library(nycflights13)
library(pander)

knitr::opts_chunk$set(echo = F, cache = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)

```

# Introduction

Here we port the 01_visualization.ipynb notebook into an Rmarkdown file to be rendered into an html document.

We can use markdown syntax:

- list items
- like that

*italic*

**bold**

***bold and italic***

Formulas using latex:

$${E=mc^{2}}$$

Or change formatting with html:

<font color="red"> Red colored tex </font>

We can cite resources in bibliography file and they will be added to the references automatically [@savitch2001; @salisbury1997]

We can add in line R code like first ten numbers: `r pander(1:10)`

We can add a chunk for a fancy table:

```{r}
kable(head(iris, 5), booktabs = TRUE) %>%
  kable_styling(font_size = 8)
```

In Rmarkdown codes are inserted by chunks or inline, there are no cells as in jupyter notebooks.

`rmarkdown::convert_ipynb` function can convert an ipynb to an Rmd file.

# ggplot2

Let's wrangle flights slightly:

```{r echo = T}
flights2 <- copy(flights)
setDT(flights2)
flights2[, date1 := as.Date(time_hour)]
flights2 <- flights2 %>% mutate_at(vars(ends_with("time")), function(x) hm(sprintf("%s:%s", x %/% 100, x %% 100)))
flights2[, speed := distance / period_to_seconds(air_time) * 3600]
```

Now first let's create a line chart across dates, where y axis shows the average departure delay for each day

The dimensions are passed by `aes()` function:

```{r}
flights2 %>%
group_by(date1) %>%
summarise_at("dep_delay", mean, na.rm = T) %>%
ggplot(aes(x = date1, y = dep_delay)) +
geom_line()
```

Now let's differentiate the lines by color according to the origin airport, so we add a third dimension:

```{r}
flights2 %>%
group_by(date1, origin) %>%
summarise_at("dep_delay", mean, na.rm = T) %>%
ggplot(aes(x = date1, y = dep_delay, color = origin)) +
geom_line()
```

Let's try a scatter plot across average daily departure and array delays:

```{r}
flights2 %>%
group_by(date1) %>%
summarise_at(c("dep_delay", "arr_delay"), mean, na.rm = T) %>%
ggplot(aes(x = dep_delay, y = arr_delay)) +
geom_point()
```

Let's add a third dimension by changing the size of points according to the flight distance, so we have a bubble chart:

```{r}
flights2 %>%
group_by(date1) %>%
summarise_at(c("dep_delay", "arr_delay", "distance"), mean, na.rm = T) %>%
ggplot(aes(x = dep_delay, y = arr_delay, size = distance)) +
geom_point(alpha = 0.5)
```

And let's add a fourth dimension by changing the color of points according to the origin airport:

```{r}
flights2 %>%
group_by(date1, origin) %>%
summarise_at(c("dep_delay", "arr_delay", "distance"), mean, na.rm = T) %>%
ggplot(aes(x = dep_delay, y = arr_delay, size = distance, color = origin)) +
geom_point(alpha = 0.5)
```

And we can create multiple charts for each weekday using `facet_wrap()` function, now we have five dimensions:

```{r}
flights2 %>%
group_by(date1, origin) %>%
summarise_at(c("dep_delay", "arr_delay", "distance"), mean, na.rm = T) %>%
mutate(weekday = lubridate::wday(date1, label = T)) %>%
ggplot(aes(x = dep_delay, y = arr_delay, size = distance, color = origin)) +
geom_point(alpha = 0.5) +
facet_wrap(. ~ weekday)
```

And let's create a separate chart for each combinations of a weekday and a quarter (three months) using `facet_grid()` function, we have six dimensions!:

```{r}
flights2 %>%
group_by(date1, origin) %>%
summarise_at(c("dep_delay", "arr_delay", "distance"), mean, na.rm = T) %>%
mutate(weekday = lubridate::wday(date1, label = T)) %>%
mutate(mnt = month(date1)) %>%
mutate(quartx = paste("Q", (mnt - 1) %/% 3 + 1), sep = "") %>%
ggplot(aes(x = dep_delay, y = arr_delay, size = distance, color = origin)) +
geom_point(alpha = 0.5) +
facet_grid(weekday ~ quartx)
```

# plotly

So far we had static charts with no interactions, just pictures.

Now let's do something fancy very easily:

- First create a ggplot chart and assign to a named object
- Call that object with `ggplotly()` function from plotly package

```{r}
pl1 <- flights2 %>%
group_by(date1, origin) %>%
summarise_at(c("dep_delay", "arr_delay", "distance"), mean, na.rm = T) %>%
ggplot(aes(x = dep_delay, y = arr_delay, size = distance, color = origin)) +
geom_point(alpha = 0.5)

ggplotly(pl1)
```

It is the same chart, but we can hover over points and see the data in pop-up tooltip, we can switch colors on and off, zoom, pan, etc.

Faceted charts can also be converted to plotly:

```{r}
pl2 <- flights2 %>%
group_by(date1, origin) %>%
summarise_at(c("dep_delay", "arr_delay", "distance"), mean, na.rm = T) %>%
mutate(weekday = lubridate::wday(date1, label = T)) %>%
ggplot(aes(x = dep_delay, y = arr_delay, size = distance, color = origin)) +
geom_point(alpha = 0.5) +
facet_wrap(. ~ weekday)

ggplotly(pl2)
```

We can also create animated charts using plotly's own syntax and passing a feature for the `frame` dimensions:

```{r}
flights2 %>%
group_by(date1, origin) %>%
summarise_at(c("dep_delay", "arr_delay", "distance"), mean, na.rm = T) %>%
mutate(mnt = month(date1)) %>%
plot_ly(x = ~dep_delay, y = ~arr_delay) %>%
add_trace(marker = list(color = origin), frame = ~mnt, type = "scatter") %>%
animation_opts(
    frame = 200, redraw = T, easing = "linear", mode = "next"
)
```


We can even create 3d interactive rotatable charts easily in plotly:

```{r}
flights3 <- flights2 %>%
group_by(date1, origin) %>%
summarise_at(c("dep_delay", "arr_delay", "distance"), mean, na.rm = T) %>%
mutate(mnt = month(date1))

if (T)
{    
    plot_ly() %>% 
          add_trace(data = flights3,  x = flights3$dep_delay, y = flights3$arr_delay, z = flights3$distance, type="mesh3d") %>%
            layout(autosize = F, width = 400, height = 400,
                  scene = list(xaxis = list(title = "dep_delay"),
                  yaxis = list(title = "arr_delay"),
                  zaxis = list(title = "distance")))
}
```

# Conclusion

So we can render any interactive plots in an html document format

# References
